---
title: "Quick Reminder"
author: "Alejandra Medina-Rivera"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
---



```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment = "")
```

#Access R

* R can be accessed through the command line 

```{r, include=FALSE,  eval=FALSE, echo=TRUE, warning=FALSE}
R
```

* Emacs has a package to communicate with R, Emacs Speaks Statistics

* R can be accessed through a Graphic User Interface
We will be using [Rstudio](https://www.rstudio.com/)

#Some basic notes
##Operators in R

Operator      | Description
------------- | -------------
+  | addition
- | subtraction
* | multiplication
/ | division
** or ^ | exponential
> | greater than
>= | greater than or equal to
== | equal to
!= | not equal to

#Data Types
In R, the basic data types are vectors, not scalars
A vector contains an index set of values that are all of the same type:
* logical 
* numeric 
* complex 
* character

The numeric type can be further broken down into integer, single, and double types (but this important when making calls to foreign functions, e.g. C or FORTRAN)

##Data Structures
The principal data structures in R are:
* vector- array of objects of the same type 
* matrix- array of vectors
* list- can contain objects of different types 
* environment- hash table 
* data.frame-array of vectors, lists or both. 
* factor- categorical
* function
Packages as [Bioconductor](https://www.bioconductor.org/) provide other types of data structures.

##Declaring variables
The operators to create variables are **<-** and **=**

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
v1 <- c(1:10)
v1
v2 <- runif(10)
v2
v3 <- sample(c("A", "C", "G", "T"),
    size = 10, replace = TRUE)
v3
v4 <- v3 %in% c("A", "G")
v4
v5 <- c("foo", 2, TRUE)
v5
v6 <- c(2, "3")
v6
```

## Install pacakges

For the sake of this course it is required to install some packages.
```{r, include=TRUE,  eval=FALSE, echo=TRUE, warning=FALSE}
install.packages("devtools", "roxygen2", "Rcpp", "RcppArmadillo", "knitr")
install.packages("ISwR")

```
##Control Structures
* if is the most simple control structure, and usage is simple:
if (cond1=vdd) {cmd1} else {cmd2}

* ifelse, usage: ifelse(test, vaue-true, value-false)

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
 if (1 == 0) {  print(1)  } else {
print(2) }

x <- 1:10
ifelse(x < 5 | x > 8, x, 0)

```
## Input data
* There are several functions in R to read in data from a file
    + scan()
    + read.table()
    + read.csv()
    + source().

* Remember, to learn more: help(read.csv)
* scan() is useful when you don???t know the structure of your data..
* source() command used to read scripts R and execute them inside the current session.

#Cycling
In R there are three cycling structures.

##Basic cycling commands
* **while** Usage: while(condition) {...} 

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
y <- 12345
x <- y/2
x
while (abs(x * x - y) > 1e-10) 
{
  x <- (x + + y/x)/2
  }
x
x^2
```


* **for** Usage: for(variable in sequence) {...} 
```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
x <- seq(0, 1, 0.05)
plot(x, x, ylab = "y", type = "l") > for (j in 2:8) {
lines(x, x^j)
}
```

## Apply
A common application of loops is to apply a function to each element of a set of values or vectors and collect the results in a single structure. 

In R this is abstracted by the family of apply functions. 
For example:
* lapply, returns a list.
* sapply, simplifies the result to a vector or a matrix if possible.

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
library(ISwR)
data( thuesen )
head(thuesen)
lapply(thuesen, mean, na.rm = T)
sapply(thuesen, mean, na.rm = T)
```
Why these and not explicit loops?

* These functions attach meaningful names to the results
* When correctly used they tend to be faster

An other function of this family is apply, which allows you to apply a function to the row or the columns of a matrix
```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
m <- matrix(rnorm(12), 4) 
m
apply(m, 2, min)
```

#Functions

* Functions are objects of type function and all functions in R as mean are of the same type.
* Functions are defined as:
mifun <- function(arg1, arg2, ...) { What the function will do}

* Functions are called with:
mifun(arg1 = ..., arg2 = ...)

* Is better if all arguments have a default value ex: arg1 = val.def.
When a function is called, the arguments can be set in the same order these are defined in the function, or you can use their tag (names)

* Inside the function there can be one or several instructions. 
* The returned value of a function is the last one to be evaluated or the one defined with return.

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
fact <- function(x = 1) {
  ret <- 1
  for (i in 1:x) {
    ret <- ret *i 
  }
  return(c(x,ret))
}
fact()
fact(x=5)
fact(6)
```

## Control
* In R there are three control functions that can be used within function to regulate how they work

    + **return**: specifies the value that has to be returned and ends the function.
    + **stop**: stops the function and prints an error message.
    + **warning**: prints a message but doesn't stop the function.

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
myfun <- function(x1){
  if (x1 > 0) {
    print (x1)
    }
else if (x1 == 0) {
  warning("Value must be > 0")
  }
else{
  stop ("There is a problem, x1 < 0 ")
  }
}
```
```{r, include=TRUE,  eval=FALSE, echo=TRUE, warning=FALSE}

myfun(x1=0)
myfun(x1=2)

```

## References
* [Introductory Statistics with R,  Dalgaard, Peter, Springer]( https://www.springer.com/us/book/9780387790534)


