---
title: "Quick Reminder"
author: "Alejandra Medina-Rivera"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
---



```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment = "")
```

#Access R

* R can be accessed through the command line 

```{r, include=FALSE,  eval=FALSE, echo=TRUE, warning=FALSE}
R
```

* Emacs has a pacakge to communicate with R, Emacs Speaks Statistics

* R can be accessed through a Graphic User Interface
We will be using [Rstudio](https://www.rstudio.com/)

#Some basic notes
##Operators in R

Operator      | Description
------------- | -------------
+  | addition
- | substraction
* | multiplication
/ | division
** or ^ | exponential
> | greater than
>= | greater than or equal to
== | equal to
!= | not equal to

#Data Types
In R, the basic data types are vectors, not scalars
A vector contains an index set of values that are all of the same type:
* logical 
* numeric 
* complex 
* character

The numeric type can be further broken down into integer, single, and double types (but this important when making calls to foreign functions, eg. C or Fortran)

##Data Structures
The principal data structures in R are:
* vector- array of objects of the same type 
* matrix- array of vectors
* list- can contain objects of different types 
* environment- hashtable 
* data.frame-array of vectors, lists or both. 
* factor- categorical
* function
Packages as [Bioconductor](https://www.bioconductor.org/) provide other types of data structures.

##Declaring variables
The operators to create variables are **<-** and **=**

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
v1 <- c(1:10)
v1
v2 <- runif(10)
v2
v3 <- sample(c("A", "C", "G", "T"),
    size = 10, replace = TRUE)
v3
v4 <- v3 %in% c("A", "G")
v4
v5 <- c("foo", 2, TRUE)
v5
v6 <- c(2, "3")
v6
```

##Control Structures
* if is the most simple control structure, and usage is simple:
if (cond1=vdd) {cmd1} else {cmd2}

* ifelse, usage: ifelse(prueba, valor-vdd, valor-falso)

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
 if (1 == 0) {  print(1)  } else {
print(2) }

x <- 1:10
ifelse(x < 5 | x > 8, x, 0)

```
## Input data
* There are several functions in R to read in data from a file
    + scan()
    + read.table()
    + read.csv()
    + source().

* Remember, to learn more: help(read.csv)
* scan() is useful when you don???t know the structure of your data..
* source() command used to read scripts R and execute them inside the current session.

#Cycling
In R there are three cycling structures.

* **while** Usage: while(condition) {...} 

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
y <- 12345
x <- y/2
x
while (abs(x * x - y) > 1e-10) 
{
  x <- (x + + y/x)/2
  }
x
x^2
```


* **for** Usage: for(variable in sequence) {...} 
```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
x <- seq(0, 1, 0.05)
plot(x, x, ylab = "y", type = "l") > for (j in 2:8) {
lines(x, x^j)
}
```


#Functions

* Functions are objects of type function and all functions in R as mean are of the same type.
* Functions are defined as:
mifun <- function(arg1, arg2, ...) { What the function will do}

* Functions are called with:
mifun(arg1 = ..., arg2 = ...)

* Is better if all arguments have a default value ex: arg1 = val.def.
When a function is called, the arguments can be set in the same order these are defined in the function, or you can use their tag (names)

* Inside the function there can be one or several instructions. 
* The returned value of a function is the last one to be evaluated or the one defined with return.

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
fact <- function(x = 1) {
  ret <- 1
  for (i in 1:x) {
    ret <- ret *i 
  }
  return(c(x,ret))
}
fact()
fact(x=5)
fact(6)
```

## Control
* In R there are three control functions that can be used within function to regulate how they work

    + **return**: specifies the value that has to be returned and ends the function.
    + **stop**: stops the function and prints an error message.
    + **warning**: prints a message but doesn't stop the function.

```{r, include=TRUE,  eval=TRUE, echo=TRUE, warning=FALSE}
myfun <- function(x1){
  if (x1 > 0) {
    print (x1)
    }
else if (x1 == 0) {
  warning("Value must be > 0")
  }
else{
  stop ("There is a problem, x1 < 0 ")
  }
}
```
```{r, include=TRUE,  eval=FALSE, echo=TRUE, warning=FALSE}

myfun(x1=0)
myfun(x1=2)

```

#Basics for Package Development
Packages are one of the great advantages of R, they provide an efficient way to share and mantain code and methods.
The functions can be private or shared across a comunity.
Moreover, the packakging system in R has many useful tools to validate the software, for example: To check the documetation exists and is in sync with the code.


